---
title: "Prediction Assignment Writeup"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. This task is to perform analysis on barbell lifts correctly and incorrectly in 5 different ways from the data by accelerometers on the belt, forearm, arm and dumble of 6 participants.

1. Training data: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

2. Test data: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

These data comes from http://groupware.les.inf.puc-rio.br/har.

## Loading Datasets
Download the datasets and read it into data table.
```{r loadData}
require(data.table)

url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
D <- fread(url)
```

```{r loadTestData}
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
DTest <- fread(url)
```

`belt`, `arm`, `dumbbell`, and `forearm` variables that do not have any missing values in the test dataset will be predictor candidates.
```{r selectPredictors}
isAnyMissing <- sapply(DTest, function (x) any(is.na(x) | x == ""))
isPredictor <- !isAnyMissing & grepl("belt|[^(fore)]arm|dumbbell|forearm", names(isAnyMissing))
predCandidates <- names(isAnyMissing)[isPredictor]

predCandidates
```

Subset the primary dataset to include only the predictor candidates and the outcome variable `classe`.
```{r compilePredictors}
varToInclude <- c("classe", predCandidates)
D <- D[, varToInclude, with=FALSE]

dim(D)
```

```{r viewData}
names(D)
```

* Compile `classe` into factor to be used in model.
```{r compileFactors}
D <- D[, classe := factor(D[, classe])]
D[, .N, classe]
```

* Split the dataset into a 60% training and 40% probing dataset.
```{r splitSet}
require(caret)

seed <- as.numeric(as.Date("2014-10-26"))
set.seed(seed)
inTrain <- createDataPartition(D$classe, p=0.6)
DTrain <- D[inTrain[[1]]]
DProbe <- D[-inTrain[[1]]]
```

* Preprocess the prediction variables by centering and scaling.
```{r preProcess}
X <- DTrain[, predCandidates, with=FALSE]
preProc <- preProcess(X)
preProc
```

```{r centerScale}
XCS <- predict(preProc, X)
DTrainCS <- data.table(data.frame(classe = DTrain[, classe], XCS))
```

* Apply process centering and scaling to probing dataset.
```{r centerScaleProbe}
X <- DProbe[, predCandidates, with=FALSE]
XCS <- predict(preProc, X)
DProbeCS <- data.table(data.frame(classe = DProbe[, classe], XCS))
```

* Evaluate for near zero variance.
```{r zeroVar}
nzv <- nearZeroVar(DTrainCS, saveMetrics=TRUE)

if (any(nzv$nzv)) nzv else message("No variables with near zero variance")
```
//------------------------------------------------------
## Training Prediction Model
1. Setting up parallel clusters.
```{r paraCluster}
require(parallel)
require(doParallel)

clus <- makeCluster(detectCores() - 1)
registerDoParallel(clus)
```

2. Setting up control parameters.
```{r paramCtrl}
ctrl <- trainControl(classProbs=TRUE,
                     savePredictions=TRUE,
                     allowParallel=TRUE)
```

3. Fit the parameters into the model.
```{r paramFit}
method <- "rf"
system.time(trainingModel <- train(classe ~ ., data=DTrainCS, method=method))
```

4. Reset the clusters values.
```{r stopClu}
stopCluster(clus)
```

### Test the model on training dataset.
```{r loadTrainModel}
trainingModel
```

```{r testTrain}
hat <- predict(trainingModel, DTrainCS)
confusionMatrix(hat, DTrain[, classe])
```

### Test the model on test probing dataset.
```{r testProbe}
hat <- predict(trainingModel, DProbeCS)
confusionMatrix(hat, DProbeCS[, classe])
```

### Present final model
```{r varImpTrainModel}
varImp(trainingModel)
```

```{r implementFinal}
trainingModel$finalModel
```

The estimated error rate was less than 1%.
Pack the training model into RData object to be used on prediction.
```{r saveModel}
save(trainingModel, file="trainingModel.RData")
```

## Prediction on test dataset.

Load training model from previous session.
```{r loadTrainTest}
load(file="trainingModel.RData", verbose=TRUE)
```

Apply the training model to predict into the test dataset.
```{r predictEval}
DTestCS <- predict(preProc, DTest[, predCandidates, with=FALSE])
hat <- predict(trainingModel, DTestCS)
DTest <- cbind(hat , DTest)

subset(DTest, select=names(DTest)[grep("belt|[^(fore)]arm|dumbbell|forearm", names(DTest), invert=TRUE)])
```